---
title: "STAT 847: Analysis Assignment 1"
author: |
  | Andrew Girgis
  | 21108082
output: pdf_document
---

```{r setup, include=FALSE}
library(knitr)
#opts_chunk$set(tidy.opts=list(width.cutoff=70),tidy=TRUE)
knitr::opts_chunk$set(echo = TRUE)
```

## Importing Libraries
```{r, message=FALSE, warning=FALSE, include = TRUE}
library(readr)
library(formatR)
library(plyr)
library(dplyr)
library(ggplot2)
library(tidyverse)
```


```{r, include=FALSE}
#set wd
setwd('/Users/andrew/Downloads/UW courses/STAT 847/Analysis 1')
```


## Importing data

```{r}
asba_data = read.csv('HRN assiniboia scraped data.csv')
hast_data = read.csv('HRN hastings scraped data.csv')
wdbn_data = read.csv('HRN woodbine scraped data.csv')
```

```{r, echo = FALSE, results='hide'}
print(head(asba_data))
```

```{r, echo=FALSE, results='hide'}
print(head(hast_data))
```

```{r, echo=FALSE,results='hide'}
print(head(wdbn_data))
```

```{r}
colnames(wdbn_data)
```
The columns that pertain to the horse are: horse_number, horse_name,horse_sire, horse_trainer, horse_jockey, horse_odds, horse_odds_decimal,and horse_place.

Therefore the columns that dont pertain to the horse are:meet_location, meet_wday, meet_mday, meet_year, racecount, race_number, purse, time_frac1, time_frac2, time_frac3, time_frac4, time_frac5, time_final, track_length, track_type, race_class, dist_frac1, dist_frac2, dist_frac3, dist_frac4, dist_frac5

```{r, include=FALSE}
str(wdbn_data)
```
\newpage

## Question 1
```{r}
#Transposing our data so that each row represents a race grouping by the race count
race_stats = ddply(wdbn_data, .(racecount), summarize,
                   meet_location = meet_location[1], 
                   meet_wday = meet_wday[1], 
                   meet_mday = meet_mday[1], 
                   meet_year = meet_year[1], 
                   race_number = race_number[1], 
                   purse = purse[1], 
                   time_frac1 = time_frac1[1], 
                   time_frac2 = time_frac2[1], 
                   time_frac3 = time_frac3[1], 
                   time_frac4 = time_frac4[1] , 
                   time_frac5 = time_frac5[1],
                   time_final = time_final[1], 
                   track_length = track_length[1] , 
                   track_type = track_type[1], 
                   race_class = race_class[1], 
                   dist_frac1 = dist_frac1[1], 
                   dist_frac2 = dist_frac2[1], 
                   dist_frac3 = dist_frac3[1], 
                   dist_frac4 = dist_frac4[1], 
                   dist_frac5 = dist_frac5[1]
                   )

#prints the first 3 rows of the dataframe
print(head(race_stats, 3))

```
\newpage

## Question 2

```{r, tidy=TRUE}
#using the by() function to calculate the average time it takes for the winning horse to complete a race of each available length 
time_by_length = by(race_stats$time_final, race_stats$track_length, mean)

# creating a variable track_length with all available track lengths
track_length <- c("1 1/16M", "1 1/2M", "1 1/4M", "1 1/8M", "1 3/4M", "1 3/8M", "1M", "1M 70Y", "4 1/2F", "5 1/2F", "5F", "6 1/2F", "6F", "7 1/2F", "7F")

#creating a variable average_time that contains all the averages tiems per track length
average_time <- c(86.20903, 91.73857, 87.32706, 84.53562, 92.31, 88.09571, 88.01048, 98.694, 65.1875, 77.8134, 77.02262, 82.01258, 82.86664, 84.23183, 84.0166)

#combining track_length and average_time into a dataframe
track_data <- data.frame(track_length, round(average_time, 2))


```


\begin{table}[h]
  \centering
  \begin{tabular}{lr}
    \hline
    \hline
    Event length & Average Time \\
    \hline
    4 1/2F (4.5 Furlongs) & 65.19 \\
    5F (5 Furlongs) & 77.02 \\
    5 1/2F (5.5 Furlongs) & 77.81 \\
    6F (6 Furlongs) & 82.87 \\
    6 1/2F (6.5 Furlongs) & 82.01 \\
    7F (7 Furlongs) & 84.02 \\
    7 1/2F (7.5 Furlongs) & 84.23 \\
    1M (1 Mile, 8 Furlongs) & 88.01 \\
    1 1/16M & 86.21 \\
    1 1/8M (1.125 Miles) & 84.54 \\
    1 1/4M (1.25 Miles) & 87.33 \\
    1 3/8M & 88.10 \\
    1 1/2M (1.5 Miles, 12 Furlongs) & 91.74 \\
    1 3/4M (1.75 Miles) & 92.31 \\
    1M 70Y & 98.69 \\
    \hline
    \hline
  \end{tabular}
  \caption{Table depicting average time to complete a race to complete a race of each length}
\end{table}

\newpage

## Question 3

```{r, results='hide', tidy=TRUE}
wdbn_data$floor_odds = round(wdbn_data$horse_odds_decimal)

tab1 = table(wdbn_data$floor_odds, wdbn_data$horse_place)
tab1

tab2 = round(prop.table(tab1, 1),3)
tab2

```


\begin{table}[h]
  \centering
  \begin{tabular}{lr}
    \hline
    \hline
    Rounded Odds & Probability of 2nd \\
    \hline
    0 & 0.500 \\
    1 & 0.250 \\
    2 & 0.216 \\
    3 & 0.196 \\
    4 & 0.147 \\
    5 & 0.120 \\
    6 & 0.111 \\
    8 & 0.107 \\
    10 & 0.094 \\
    12 & 0.099 \\
    15 & 0.065 \\
    20 & 0.051 \\
    30 & 0.000 \\
    50 & 0.000 \\
    \hline
    \hline
  \end{tabular}
  \caption{Table depicting the probability of a horse coming in second place as a function of the decimal odds.}
\end{table}

\newpage

## Question 4

```{r, results='hide', tidy=TRUE}
wdbn_6f = subset(wdbn_data, wdbn_data$track_length=="6F")
turf_data <- subset(wdbn_6f, track_type %in% c('Inner turf', 'Turf'))
all_weather_data <- subset(wdbn_6f, track_type == 'All Weather Track')

# Perform the two-sample t-test
t_test_result <- t.test(turf_data$time_final, all_weather_data$time_final)

tstat <- t_test_result$statistic
pval <- t_test_result$p.value

alpha <- 0.05
if (t_test_result$p.value < alpha) {
  cat('Reject the null hypothesis. There is a significant difference in average finish times.\n')
} else {
  cat('Fail to reject the null hypothesis. There is no significant difference in average finish times.\n')
}
```
T-Statistic: `r tstat`

P-Value: `r pval`

Fail to reject the null hypothesis. There is no significant difference in average finish times.

\newpage

## Question 5

```{r, tidy=TRUE}
asba_6f = subset(asba_data, asba_data$track_length=="6F")
hast_6f = subset(hast_data, hast_data$track_length=="6F")


wdbn_box <- ggplot(wdbn_6f, aes(x=track_length, y=time_final)) +
  geom_boxplot(outlier.colour="red", outlier.shape=8,
             outlier.size=2) +
  scale_y_continuous(name="Finish times (s)",breaks=seq(0,200,25)) +
  labs(x = "Track length (Furlongs) ") +
  ggtitle("Boxplot of track length on finish times at Woodbine") +
  theme(plot.title = element_text(hjust = 0.5))

asba_box <- ggplot(asba_6f, aes(x=track_length, y=time_final)) +
  geom_boxplot(outlier.colour="red", outlier.shape=8,
             outlier.size=2) +
  labs(y= "Finish times (s)", x = "Track length (Furlongs)") +
  ggtitle("Boxplot of track length on finish times at Assianboa") +
  theme(plot.title = element_text(hjust = 0.5))

hast_box <- ggplot(hast_6f, aes(x=track_length, y=time_final)) +
  geom_boxplot(outlier.colour="red", outlier.shape=8,
             outlier.size=2) +
  labs(y= "Finish times (s)", x = "Track length (Furlongs)") +
  ggtitle("Boxplot of track length on finish times at Hastings Park") +
  theme(plot.title = element_text(hjust = 0.5))

```

```{r, figures-side, fig.show="hold", out.width="33%", echo=FALSE, fig.cap="Boxplot of the finish times for 6F races between the three locations."}
par(mar = c(1, 3, .1, .1))
wdbn_box
asba_box
hast_box
```

Note: The red star indictates outliers in the data.

\newpage

## Question 6

```{r, message=FALSE, results='hide', tidy=TRUE}
#use min and max year to ensure we are only looking at data between 2022 and 2023
min_yr = min(wdbn_data$meet_year)
max_yr = max(wdbn_data$meet_year)

place_tab = table(wdbn_data$horse_name, wdbn_data$horse_place)
place_tab

sorted_tab <- place_tab %>% 
  as.data.frame() %>% 
  arrange(desc(Freq))
sorted_tab

sorted_tab = subset(sorted_tab, sorted_tab$Var2==1)
```

\begin{table}[h]
  \centering
  \begin{tabular}{lr}
    \hline
    \hline
    Horse & Wins \\
    \hline
    Canadiansweetheart & 8 \\
    Patches O'Houlihan & 8 \\
    Hallie's Hero & 6 \\
    Wentru & 6 \\
    C C's Kingdom & 5 \\
    \hline
    \hline
  \end{tabular}
  \caption{Table depicting the five horses that have won the most events at Woodbine}
\end{table}

\newpage


## Question 7

```{r, results='hide', tidy=TRUE}
#Find the min and max horse place
min_place = min(wdbn_data$horse_place, na.rm = TRUE)
min_place
max_place = max(wdbn_data$horse_place, na.rm = TRUE)
max_place
# Since we only have horse place from 1 to 5 we cant use that for the last 10% of
#the earnings however we can use it for the top 3 purse distributions.
```


```{r, include = FALSE}
print("Position of missing values:")
which(is.na(wdbn_data))

sum(is.na(wdbn_data$horse_place))
```

```{r, message=FALSE, results='hide', warning=FALSE, tidy=TRUE}

sub_wdbn_data <-subset(wdbn_data, (!is.na(wdbn_data[,14])))

num_horses = count(sub_wdbn_data, racecount)


sub_wdbn_data <- merge(sub_wdbn_data,num_horses,by="racecount")
      
for (row in 1:nrow(sub_wdbn_data)){
  sub_wdbn_data$payout[row] = 0
  if (isTRUE(sub_wdbn_data$horse_place[row] == 1)==TRUE){
   sub_wdbn_data$payout[row] = sub_wdbn_data$purse[row] * 0.6
  } else if (isTRUE(sub_wdbn_data$horse_place[row] == 2)==TRUE){
    sub_wdbn_data$payout[row] = sub_wdbn_data$purse[row] * 0.2
  } else if (isTRUE(sub_wdbn_data$horse_place[row] == 3)==TRUE){
    sub_wdbn_data$payout[row] = sub_wdbn_data$purse[row] * 0.1
  } else {
    sub_wdbn_data$payout[row] = (wdbn_data$purse[row] * 0.1)/(sub_wdbn_data$n - 3)
    #remember to divide by (number of horses - 3)
  }
}

total_pay = sub_wdbn_data %>%
              group_by(horse_name) %>% 
              summarise(num = n(),
                        totalpayout = sum(payout))

total_pay <- total_pay[order(total_pay$totalpayout, decreasing = TRUE),]

```

\begin{table}[h]
  \centering
  \begin{tabular}{lr}
    \hline
    \hline
    Horse & Prize Money (in \$) \\
    \hline
    Last Call & 918,802 \\
    Patches O'Houlihan & 698,862 \\
    Malibu Secret & 690,930 \\
    Bushido & 686,750 \\
    Moira & 675,502 \\
    \hline
    \hline
  \end{tabular}
  \caption{Table depicting the five horses that have won the most prize money at Woodbine}
\end{table}


\newpage

## Question 8

```{r, echo=FALSE, results='hide', tidy=TRUE}
unique(race_stats$track_length)

q8_df <- subset(race_stats, track_length != "1 3/4M")

unique(q8_df$track_length)

for (row in 1:nrow(q8_df)) {
  if (isTRUE(q8_df$track_length[row] == "4 1/2F")==TRUE) {
    q8_df$furlongs[row] = 4.5
  } else if (isTRUE(q8_df$track_length[row] == "5F")==TRUE) {
    q8_df$furlongs[row] = 5
  } else if (isTRUE(q8_df$track_length[row] == "6F")==TRUE) {
    q8_df$furlongs[row] = 6
  } else if (isTRUE(q8_df$track_length[row] == "6 1/2F")==TRUE) {
    q8_df$furlongs[row] = 6.5
  } else if (isTRUE(q8_df$track_length[row] == "7F")==TRUE) {
    q8_df$furlongs[row] = 7
  } else if (isTRUE(q8_df$track_length[row] == "7 1/2F")==TRUE) {
    q8_df$furlongs[row] = 7.5
  } else if (isTRUE(q8_df$track_length[row] == "1M")==TRUE) {
    q8_df$furlongs[row] = 8
  } else if (isTRUE(q8_df$track_length[row] == "1M 70Y")==TRUE) {
    q8_df$furlongs[row] = 8.318
  } else if (isTRUE(q8_df$track_length[row] == "1 1/16M")==TRUE) {
    q8_df$furlongs[row] = 8.5
  } else if (isTRUE(q8_df$track_length[row] == "1 1/8M")==TRUE) {
    q8_df$furlongs[row] = 9
  } else if (isTRUE(q8_df$track_length[row] == "1 1/4M")==TRUE) {
    q8_df$furlongs[row] = 10
  } else if (isTRUE(q8_df$track_length[row] == "1 3/8M")==TRUE) {
    q8_df$furlongs[row] = 11.5
  } else if (isTRUE(q8_df$track_length[row] == "1 1/2M")==TRUE) {
    q8_df$furlongs[row] = 12
  } else if (isTRUE(q8_df$track_length[row] == "1 3/4M")==TRUE) {
    q8_df$furlongs[row] = 13.75
  }
}
summary(q8_df$furlongs)

tapply(q8_df$time_frac2, q8_df$furlongs, mean) 
#I am getting NA values using this method so i must check the data for NA in timefrac2

sum(is.na(q8_df$time_frac2))
#We do have NAs!

which(is.na(q8_df$time_frac2))
#now that we have located them and we only have 3/1710(1.75%) NA in the data I find it more useful to drop the rows rather than fill since they wont benfit the outcome of the mean

q8_df[c(1037,1512,1557), ]
#just to check the rows and ensure we are dropping the right ones

q8_df <- q8_df[-c(1037,1512,1557), ] 

testin = tapply(q8_df$time_frac2, q8_df$furlongs, mean)

table_q8 = data.frame(sort(unique(q8_df$furlongs)), testin)

colnames(table_q8) <- c("furlongs", "mean_time2")


```

```{r, echo=FALSE, fig.cap="Broken line graph of average time the horse takes to complete 1/2 miles against distance in furlongs" }
ggplot(table_q8, aes(x = furlongs, y = mean_time2)) +
  geom_line() +   
  geom_point(col='red') +  
  labs(x = "Distance (Furlongs)", y = "Time(s)") +  
  ggtitle("Broken line graph of 1/2M time against distance") 

```
```{r, results='hide', tidy=TRUE}
unique(race_stats$track_length)

q8_df <- subset(race_stats, track_length != "1 3/4M")

unique(q8_df$track_length)

for (row in 1:nrow(q8_df)) {
  if (isTRUE(q8_df$track_length[row] == "4 1/2F")==TRUE) {
    q8_df$furlongs[row] = 4.5
  } else if (isTRUE(q8_df$track_length[row] == "5F")==TRUE) {
    q8_df$furlongs[row] = 5
  } else if (isTRUE(q8_df$track_length[row] == "6F")==TRUE) {
    q8_df$furlongs[row] = 6
  } else if (isTRUE(q8_df$track_length[row] == "6 1/2F")==TRUE) {
    q8_df$furlongs[row] = 6.5
  } else if (isTRUE(q8_df$track_length[row] == "7F")==TRUE) {
    q8_df$furlongs[row] = 7
  } else if (isTRUE(q8_df$track_length[row] == "7 1/2F")==TRUE) {
    q8_df$furlongs[row] = 7.5
  } else if (isTRUE(q8_df$track_length[row] == "1M")==TRUE) {
    q8_df$furlongs[row] = 8
  } else if (isTRUE(q8_df$track_length[row] == "1M 70Y")==TRUE) {
    q8_df$furlongs[row] = 8.318
  } else if (isTRUE(q8_df$track_length[row] == "1 1/16M")==TRUE) {
    q8_df$furlongs[row] = 8.5
  } else if (isTRUE(q8_df$track_length[row] == "1 1/8M")==TRUE) {
    q8_df$furlongs[row] = 9
  } else if (isTRUE(q8_df$track_length[row] == "1 1/4M")==TRUE) {
    q8_df$furlongs[row] = 10
  } else if (isTRUE(q8_df$track_length[row] == "1 3/8M")==TRUE) {
    q8_df$furlongs[row] = 11.5
  } else if (isTRUE(q8_df$track_length[row] == "1 1/2M")==TRUE) {
    q8_df$furlongs[row] = 12
  } else if (isTRUE(q8_df$track_length[row] == "1 3/4M")==TRUE) {
    q8_df$furlongs[row] = 13.75
  }
}
summary(q8_df$furlongs)

tapply(q8_df$time_frac2, q8_df$furlongs, mean) 
#I am getting NA values using this method so i must check the data for NA in timefrac2

sum(is.na(q8_df$time_frac2))
#We do have NAs!

which(is.na(q8_df$time_frac2))
#now that we have located them and we only have 3/1710(1.75%) NA in the data I find it more useful to drop the rows rather than fill since they wont benfit the outcome of the mean

q8_df[c(1037,1512,1557), ]
#just to check the rows and ensure we are dropping the right ones

q8_df <- q8_df[-c(1037,1512,1557), ] 

testin = tapply(q8_df$time_frac2, q8_df$furlongs, mean)

table_q8 = data.frame(sort(unique(q8_df$furlongs)), testin)

colnames(table_q8) <- c("furlongs", "mean_time2")


```

```{r, results='hide', fig.keep='none', fig.cap="Broken line graph of average time the horse takes to complete 1/2 miles against distance in furlongs" }
ggplot(table_q8, aes(x = furlongs, y = mean_time2)) +
  geom_line() + 
  geom_point(col='red') + 
  labs(x = "Distance (Furlongs)", y = "Time(s)") +
  ggtitle("Broken line graph of 1/2M time against distance") 

```

\newpage

## Question 9

```{r, include=FALSE, tidy=TRUE}

for (row in 1:nrow(race_stats)) {
  if (isTRUE(race_stats$track_length[row] == "4 1/2F")==TRUE) {
    race_stats$furlongs[row] = 4.5
  } else if (isTRUE(race_stats$track_length[row] == "5F")==TRUE) {
    race_stats$furlongs[row] = 5
  } else if (isTRUE(race_stats$track_length[row] == "6F")==TRUE) {
    race_stats$furlongs[row] = 6
  } else if (isTRUE(race_stats$track_length[row] == "6 1/2F")==TRUE) {
    race_stats$furlongs[row] = 6.5
  } else if (isTRUE(race_stats$track_length[row] == "7F")==TRUE) {
    race_stats$furlongs[row] = 7
  } else if (isTRUE(race_stats$track_length[row] == "7 1/2F")==TRUE) {
    race_stats$furlongs[row] = 7.5
  } else if (isTRUE(race_stats$track_length[row] == "1M")==TRUE) {
    race_stats$furlongs[row] = 8
  } else if (isTRUE(race_stats$track_length[row] == "1M 70Y")==TRUE) {
    race_stats$furlongs[row] = 8.318
  } else if (isTRUE(race_stats$track_length[row] == "1 1/16M")==TRUE) {
    race_stats$furlongs[row] = 8.5
  } else if (isTRUE(race_stats$track_length[row] == "1 1/8M")==TRUE) {
    race_stats$furlongs[row] = 9
  } else if (isTRUE(race_stats$track_length[row] == "1 1/4M")==TRUE) {
    race_stats$furlongs[row] = 10
  } else if (isTRUE(race_stats$track_length[row] == "1 3/8M")==TRUE) {
    race_stats$furlongs[row] = 11.5
  } else if (isTRUE(race_stats$track_length[row] == "1 1/2M")==TRUE) {
    race_stats$furlongs[row] = 12
  } else if (isTRUE(race_stats$track_length[row] == "1 3/4M")==TRUE) {
    race_stats$furlongs[row] = 13.75
  }
}
```

```{r, tidy=TRUE}
model = lm(q8_df$time_frac2 ~ I(q8_df$furlongs^2) + q8_df$furlongs)
summary(model)
```

